Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-11 3.Thesis/Writeup/2.Manuscript.Rmd
All but last section DONE

These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied [@zhu15; @buur18]. There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@taka17]. Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murr18]. Therefore, algorithmic convergence should be monitored carefully, **but this is not straightforward**. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence [@gelm13]. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09]. Several non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms. 
<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

In this paper we study different methods for assessing non-convergence in iterative imputation algorithms. We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications. We then address the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in `R` [@R]. For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R`. The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged. And additionally, to evaluate the behavior and performance of several diagnostic methods to identify non-convergence. With that, we formulate an informed advice on when it is safe to conclude that the algorithm is converged *enough* for valid inferences. We translate the results of the study into guidelines for applied researchers.  **ADD that these guidelines may aid to draw valid inference from incomplete data?]**

@@

Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing. The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the proportion of incomplete cases $p_{\rm inc}$ in dataset $y$. The collection of observed data values in $y$ is denoted by $y_{\rm obs}$; the missing part of $y$ is referred to as $y_{\rm mis}$. For each datapoint in $y_{\rm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$) in the imputation algorithm. The state space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the **imputed values**). The collection of $\theta$-values between $t=1$ (summarizing the state space of the algorithm at initialization) and $t=T$ (summarizing the state space for the imputed values) will be referred to as an 'imputation chain'. 
<!-- **[If not defined elsewhere, add: $\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).] ** -->

@@

In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence scenario, there is a lot of trending and some chains do not intermingle. Importantly, the chain means at the last iteration (the imputed value per imputation $\ell$) are very different between the two scenarios. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. It is obvious that non-convergence impacts the distribution of the imputed values in this example, $\dot y_{\rm imp, \ell}$. This effect may translate into the distribution of the completed data {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$ and the pooled estimate $\bar{Q}$. $\bar{Q}$ is then a  biased, invalid estimate of $Q$, since systematically higher imputed values lead to bias, and under-estimation of the variance within imputation chains poses a threat to confidence-validity. Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms. **[ADD t and l to each theta??]**

```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence. Depicted are chain means of variable $j$ in each imputation ($m=5$; $T=10$). Please note that any difference between the two scenarios (see e.g., the magnitude of the chain means on the y-axes) stems from the imputation algorithms---the two scenarios have equal incomplete data $y$."}
par(mar = c(2, 2, 0.1, 0.1))
con_plot
non_plot
```

@@

Identifying non-convergence by inspecting traceplots may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18, $\S$ 6.5.2]. Moreover, the typical traceplots to inspect only **consider** univariate $\theta$s. Monitoring univariate summaries of the state space of the algorithm may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of the imputations. Ideally, we would monitor a multivariate $\theta$.

@@

Each panel in Figure \ref{fig:diagnostics} depicts one method to identify non-convergence, applied to the two scenarios from Figure \ref{fig:non-conv}. Panel A consists of the two traceplots that may be evaluated through visual inspection. Panel B shows two versions of the $AC$: the default calculation with `R` function `stats::acf()` [@R], and manual calculation as the correlation between $\theta$ in iteration $t$ and $\theta$ in iteration $t+1$. Panel C displays the traditional computation of $\widehat{R}$ conform @gelm92, and in panel D we see $\widehat{R}$ as computed by implementing @veht19 's recommendations. **[Add labels for the two types of Rhat to the figure!]**

```{r diagnostics, fig.cap="Convergence diagnostics applied on the imputation algorithms of Figure 2. $\\theta$ = chain mean in $ \\dot y_{\\rm imp, \\ell} $."}
# load("./images/diagnostics_plot.Rdata") #bquote(theta ~ "= chain mean in "~ y[imp][",m"])
(theta + plot_layout(guides = "collect", ncol = 2)) + (ac_both + plot_layout(guides = "keep")) + old_rhat + new_rhat +  plot_annotation(tag_levels = "A", tag_suffix = ".") # add 'ncol=1' for plots under each other
```

**[REPHRASE as active about panel A]** From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around $t=2$), but intermingled nicely overall. Additionally, there was very little trending. The algorithm with pathological non-convergence showed severe non-mixing, although this gradually improved (from $t=7$). In this scenario, there was a lot of trending as well initially (up-to $t=6$), after which the chains reached a somewhat more stationary state.

@@

**Describe panel C here and why it is not optimal: De grotere variantie overall zorgt er kennelijk voor dat $\widehat{R}$ groeit. De $\widehat{R}$ geeft de foutieve boodschap dat convergentie steeds slechter gaat. Verklaring: de overdispersie assumptie klopt niet als er trend is.]** The $\widehat{R}$-values in panel D are conform our expectations. The adapted version of $\widehat{R}$ proposed by @veht19 does indeed indicate less signs of non-convergence as the number of iterations goes up.  **[explain the dip in Rhat values at t=2. Namely, because we can only use 2 of the 3 tricks by @veht19 if the number of iterations is very low (t<4). That's why the $\widehat{R}$s are more similar to the traditional GR.]** \newline
<!-- However, the superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.  -->

@@

The incompleteness conditions are chosen to reflect the difficulty of the missingness problem. The underlying assumption is that a low proportion of incomplete cases $p_{\rm inc}$ leads to quick algorithmic convergence, since there is a lot of information in the observed data. Higher $p_{\rm inc}$ yields slower convergence---unless there is so little information in $y_{\rm obs}$ that this is outweighed by the random component in the imputation algorithm. Then, convergence to a stable but highly variable state may be reached instantly. **Still, we expect that the incompleteness conditions will result in more signs of non-convergence.**

@@

Our results show the performance of the MICE algorithm under different conditions of incompleteness and early stopping. For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias. For univariate scientific estimands ($Q=\mu$ and $Q=\sigma$) we observe the largest bias in the outcome variable $Y$. And for the regression coefficients ($Q=\beta$) the bias is most pronounced in $\beta_1$ (the effect of $X_1$ on $Y$). Since there is just one estimate for $Q=r^2$ to consider, we **end up with** $Q=\mu_Y, \sigma_Y,r^2, \beta_1$. In the figures, we present all incompleteness conditions, but only early stopping conditions where $1 \leq T\leq50$. That is, the results are more or less stable for conditions where $T \geq 30$. Full results are available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/).

@@

```{r Qs, fig.height = 6.8, fig.cap="Impact of non-convergence on statistical inferences. Depicted are the bias, coverage rate (CR) and confidence interval width (CIW) of the estimates $\\bar{Q}$. We present the subset of $Q$s with the worst performance in terms of bias: the mean and standard deviation of $Y$, and the regression coefficient of $X_1$."}
mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_layout(ncol = 2, guides = "collect")
```

Figure \ref{fig:Qs} displays the influence of incompleteness and early stopping on the estimated quantities of scientific interest, $Q$s. **[Add panel labels, describe panels]** Simulation conditions that are impacted by non-convergence portray more extreme bias in the estimated $Q$s, and non-nominal coverages.

### $Q=\mu$.
The estimated univariate means seem unaffected by early stopping. This implies that approximately unbiased estimates may be obtained with as little as one iteration ($T \geq 1$). The bias in $\bar{Q}$ depends solely on the proportion of incomplete cases, with $p_{\rm inc}\geq.75$ leading to more extreme biases than other conditions. Completely unbiased estimates are only obtained in conditions where $p_{\rm inc}\leq.50$. **This is curious, since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation.** Note however, that the magnitude of the bias in these conditions is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$). We, therefore, conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.

@@

### $Q=r^2$.
In contrast to the univariate estimates, the estimated coefficient of determination is clearly affected by early stopping. That is, the bias in $Q$ does not only follow the incremental order of the incompleteness conditions, but also decreases with each additional iteration. The number of iterations necessary to reach stable, non-improving estimates depends on $p_{\rm inc}$. In conditions where $p_{\rm inc}=.05$, estimates are unbiased after one iteration. The highest number of iterations necessary to reach approximate unbiasedness is seven ($p_{\rm inc}=.95$; $T\geq7$). Completely unbiased estimates are only obtained in conditions where $p_{\rm inc}\leq.25$. This implies that even for biased estimates, $T\geq7$ would suffice to reach a stable solution.

@@

Figure \ref{fig:rhats-acs} displays results for the eight sets of diagnostic methods. The two columns in the figure represent the  non-convergence diagnostics $\widehat{R}$ and $AC$; the four rows depict the four scalar summaries $\theta$ under consideration. **[describe panels??]** We evaluate the performance of the methods by establishing whether they correctly identify conditions in which $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$. As we concluded in the last section, substantially biased estimates were only obtained in conditions where $T\leq6$, and we observed non-nominal coverage only when $T\leq3$. However, there was some persistent bias in conditions where $p_{\rm inc}\geq.75$, irrespective of the number of iterations. If the diagnostic methods are appropriate for MI, we will detect non-convergence in these conditions.

@@

```{r rhats-acs, fig.height = 6.8, fig.cap="Non-convergence diagnostics. [Split up into 2 or 4 figures??]"}
mean_Rh + mean_AC + var_Rh + var_AC + est_Rh + est_AC + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 2)
```

@@

The number of iterations necessary to obtain approximately unbiased, confidence-valid estimates corresponds to the $\widehat{R}$ threshold 1.2. The thresholds 1.1 and 1.01 seem too strict compared to the validity of the inferences.

The threshold to diagnose significant $AC$-values does not appear to be appropriate in at least the current set-up, and perhaps in iterative imputation in general. A better heuristic to diagnose non-stationarity with $AC$ may be through evaluation of the $AC$-values across iterations: If the values do not substantially decrease with $T$, approximate stationarity may be concluded.
<!-- **[similar to the elbow heuristic in PCA]**. -->

@@

1. First, check traceplots of the default $\theta$s (chain means and chain variances) for signs of pathological non-convergence. Adjust the imputation model if necessary.

2. Track multivariate $\theta$s over iterations, e.g., the novel $\theta$ that is  scientific model-independent. Or specify your own scalar summary of interest (see e.g., @buur18). Monitor these $\theta$s across iterations through visual inspection, or using a non-convergence diagnostic.

3. Compute non-convergence diagnostics $\widehat{R}$ and $AC$. Do not use the traditional calculation for $\widehat{R}$ [@gelm92], or the `R` function `stats::acf()` to compute $AC$ [@R]. Instead, calculate $\widehat{R}$ conform @veht19 and compute autocorrelations manually (see e.g., [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy)).

>>>>

These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied [@zhu15; @buur18]. There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@taka17]. Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murr18]. Therefore, algorithmic convergence should be monitored carefully, but this is not straightforward. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence [@gelm13]. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09]. Several non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms. 
<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

In this paper we study different methods for assessing non-convergence in iterative imputation algorithms. We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications. We then address the impact of inducing non-convergence in iterative imputation algorithms through model-based simulation in `R` [@R]. For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R`. The aim of the simulation study is to determine whether unbiased, confidence-valid inferences may be obtained if the algorithm has not (yet) converged. And additionally, to evaluate the behavior and performance of several diagnostic methods to identify non-convergence. With that, we formulate an informed advice on when it is safe to conclude that the algorithm is converged *enough* for valid inferences. We translate the results of the study into guidelines for applied researchers, which may facilitate drawing valid inferences from incomplete data.

@@

Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing. The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the proportion of incomplete cases $p_{\rm inc}$ in dataset $y$. The collection of observed data values in $y$ is denoted by $y_{\rm obs}$; the missing part of $y$ is referred to as $y_{\rm mis}$. For each datapoint in $y_{\rm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$) in the imputation algorithm. The state space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values). The collection of $\theta$-values between $t=1$ (summarizing the state space of the algorithm at initialization) and $t=T$ (summarizing the state space for the imputed values) will be referred to as an 'imputation chain'. 
<!-- **[If not defined elsewhere, add: $\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).] ** -->

@@

In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence scenario, there is a lot of trending and some chains do not intermingle. Importantly, the chain means at the last iteration (the imputed values per imputation $\ell$) are very different between the two scenarios. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. It is obvious that non-convergence impacts the distribution of the imputed values in this example, $\dot y_{\rm imp, \ell}$. This effect may translate into the distribution of the completed data {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}, which consequently affects the estimated quantities of scientific interest $\hat{Q}_{\ell}$ and the pooled estimate $\bar{Q}$. $\bar{Q}$ is then a  biased, invalid estimate of $Q$, since systematically higher imputed values lead to bias, and under-estimation of the variance within imputation chains poses a threat to confidence-validity. Therefore, it is important to reach algorithmic convergence in iterative imputation algorithms.

```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence (reproduced from van Buuren; 2018). Depicted are chain means of variable $j$ in each imputation ($m=5$; $T=10$). Please note that any difference between the two scenarios (see e.g., the magnitude of the chain means on the y-axes) stems from the imputation algorithms---the two scenarios have equal incomplete data $y$."}
par(mar = c(2, 2, 0.1, 0.1))
con_plot
non_plot
```

@@

Identifying non-convergence by inspecting traceplots may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18, $\S$ 6.5.2]. Moreover, traceplots are typically only used to inspect  univariate $\theta$s. Monitoring univariate summaries of the state space of the algorithm may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of the imputations. Ideally, we would monitor a multivariate $\theta$.

@@

Each panel in Figure \ref{fig:diagnostics} depicts one method to identify non-convergence, applied to the two scenarios from Figure \ref{fig:non-conv}. Panel A consists of the two traceplots that may be evaluated through visual inspection. Panel B shows two versions of the $AC$: the default calculation with `R` function `stats::acf()` [@R], and manual calculation as the correlation between $\theta$ in iteration $t$ and $\theta$ in iteration $t+1$. Panel C displays the traditional computation of $\widehat{R}$ conform @gelm92 (original), and in panel D we see $\widehat{R}$ as computed by implementing Vehtari et al.'s recommendations (adapted).

```{r diagnostics, fig.cap="Convergence diagnostics applied on the two scenarios from Figure 2 ($\\theta$ = chain mean of variable $j$; $m=5$; $T=10$). Depicted are the imputation chains in panel A, two versions of the autocorrelation in panel B, and two versions of $\\widehat{R}$ in panels C and D."}
# load("./images/diagnostics_plot.Rdata") #bquote(theta ~ "= chain mean in "~ y[imp][",m"])
(theta + plot_layout(guides = "collect", ncol = 2)) + (ac_both + plot_layout(guides = "keep")) + old_rhat + new_rhat +  plot_annotation(tag_levels = "A", tag_suffix = ".") # add 'ncol=1' for plots under each other
```

By visually inspecting the imputation chains in panel A, we conclude that the two scenarios show very different convergence. The typically converged algorithm initially portrayes some signs of non-mixing (around $t=2$), but intermingles nicely overall. Additionally, there is very little trending in this scenario. The algorithm with pathological non-convergence shows severe non-mixing, although this gradually improves (beyond $t=7$). In this scenario, we see a lot of trending initially (up-to $t=6$), after which the chains reached a somewhat more stationary state.

@@

In panel C, we do not see a lot of difference between the two scenarios. The $\widehat{R}$-values are even somewhat lower for the non-convergence scenario than for the typical convergence. And apparently, the large variance causes an increase in $\widehat{R}$-values. Convergence seems to worsen with every additional iteration. This paints the wrong picture. An explanation for this behavior may be that under severe trending, the assumption of over-dispersion is violated. Taken together, this version of $\widehat{R}$ is sub-optimal for assessing non-convergence in MI.

The $\widehat{R}$-values in panel D are conform our expectations. The adapted version of $\widehat{R}$ depicted here does indicate less signs of non-convergence as the number of iterations goes up. An exception to this general trend is the 'dip' around $t=3$. As diagnostic tool, this may lead to the incorrect conclusion that the non-convergence scenario has superior convergence compared to the typical scenario. We can explain this as a downside of the low number of iteration in iterative imputation: the adapted version of $\widehat{R}$ can only be completely employed if the number of iterations is at least four. Otherwise, it is not possible to perform all three transformations to the chains of $\theta$-values. For low values of $t$, the $\widehat{R}$-values conform Vehtari et al. are, therefore, more similar to the original $\widehat{R}$ by Gelman and Rubin. \newline
<!-- However, the superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.  -->

@@

The incompleteness conditions are chosen to reflect the difficulty of the missingness problem. The underlying assumption is that a low proportion of incomplete cases $p_{\rm inc}$ leads to quick algorithmic convergence, since there is a lot of information in the observed data. Higher $p_{\rm inc}$ yields slower convergence---unless there is so little information in $y_{\rm obs}$ that this is outweighed by the random component in the imputation algorithm. Then, convergence to a stable but highly variable state may be reached instantly. Still, we expect that incompleteness conditions with higher $p_{\rm inc}$ will result in more signs of non-convergence.

@@

Our results show the performance of the MICE algorithm under different conditions of incompleteness and early stopping. For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias. For univariate scientific estimands ($Q=\mu$ and $Q=\sigma$) we observe the largest bias in the outcome variable $Y$. And for the regression coefficients ($Q=\beta$) the bias is most pronounced in $\beta_1$ (the effect of $X_1$ on $Y$). Since there is just one estimate for $Q=r^2$ to evaluate, we consider $Q=\mu_Y, \sigma_Y,r^2, \beta_1$. In the figures, we present all incompleteness conditions, but only early stopping conditions where $1 \leq T\leq50$. That is, the results are more or less stable for conditions where $T \geq 30$. Full results are available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/).

@@

```{r Qs, fig.height = 6.8, fig.cap="Impact of non-convergence on statistical inferences. Depicted are the bias, coverage rate (CR) and confidence interval width (CIW) of the estimates $\\bar{Q}$. We present the subset of $Q$s with the worst performance in terms of bias: the mean and standard deviation of $Y$, and the regression coefficient of $X_1$."}
mean_bias + sd_bias + Rsq_bias + est_bias + est_cov + est_ciw + plot_layout(ncol = 2, guides = "collect") + plot_annotation(tag_levels = "A", tag_suffix = ".")
```

Figure \ref{fig:Qs} displays the influence of incompleteness and early stopping on the estimated quantities of scientific interest, $Q$s. In the first row of the figure, we see the bias in the estimated descriptive statistics ($Q=\mu_Y$ and $Q=\sigma_Y$). The second row shows the bias in the multivariate scientific estimands ($Q=r^2$ and $Q=\beta_1$). The third row consists of the coverage rate (CR) and confidence interval width (CIW) of the estimated regression coefficient ($Q=\beta_1$).

<!-- Simulation conditions that are impacted by non-convergence portray more extreme bias in the estimated $Q$s, and non-nominal coverages.  -->

### $Q=\mu$.
The estimated univariate means seem unaffected by early stopping. This implies that approximately unbiased estimates may be obtained with as little as one iteration ($T \geq 1$). The bias in $\bar{Q}$ depends solely on the proportion of incomplete cases, with $p_{\rm inc}\geq.75$ leading to more extreme biases than other conditions. Completely unbiased estimates are only obtained in conditions where $p_{\rm inc}\leq.50$. This is curious, since the MCAR missingness mechanism should yield unbiased univariate estimates without employing multiple imputation. Note however, that the magnitude of the bias in these conditions is small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$). We, therefore, conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.

@@

### $Q=r^2$.
In contrast to the univariate estimates, the estimated coefficient of determination is clearly affected by early stopping, see Figure \ref{fig:Qs}C. That is, the bias in $Q$ does not only follow the incremental order of the incompleteness conditions, but also decreases with each additional iteration. The number of iterations necessary to reach stable, non-improving estimates depends on $p_{\rm inc}$. In conditions where $p_{\rm inc}=.05$, estimates are unbiased after one iteration. The highest number of iterations necessary to reach approximate unbiasedness is seven ($p_{\rm inc}=.95$; $T\geq7$). Completely unbiased estimates are only obtained in conditions where $p_{\rm inc}\leq.25$. This implies that even for biased estimates, $T\geq7$ would suffice to reach a stable solution.

@@

Figure \ref{fig:rhats-acs} displays results of the eight sets of diagnostic methods. The two columns in the figure represent the  non-convergence diagnostics $\widehat{R}$ and $AC$; the rows depict the four scalar summaries $\theta$ under consideration. The first row in Figure \ref{fig:rhats-acs} thus shows $\widehat{R}$ and $AC$ applied on the chain means. In the second row we see the diagnostics applied on the chain variances. The third row depicts the same for $\theta=\hat{Q}$. And in the last row we see the diagnostics applied on the novel $\theta$ to consider: $\lambda_1$. We evaluate the performance of the methods by establishing whether they correctly identify conditions in which $\bar{Q}$ is *not* an unbiased, confidence-valid estimate of $Q$. As we concluded in the last section, substantially biased estimates were only obtained in conditions where $T\leq6$, and we observed non-nominal coverage only when $T\leq3$. However, there was some persistent bias in conditions where $p_{\rm inc}\geq.75$, irrespective of the number of iterations. If the diagnostic methods are appropriate for MI, we will detect non-convergence in these conditions.

@@

```{r rhats-acs, fig.height = 6.8, fig.cap="Non-convergence diagnostics $\\widehat{R}$ and $AC$ applied on several $\\theta$s. The left-hand side of the figure contains $\\widehat{R}$-values, and the right-hand side contains $AC$-values. Depicted in the rows are the scalar summaries $\\theta$: chain mean, chain variance, the quantity of scientific interest $\\hat{Q}$, and the first eigenvalue of the variance-covariance matrix $\\lambda_1$."}
mean_Rh + mean_AC + var_Rh + var_AC + est_Rh + est_AC + PCA_Rh + PCA_AC + plot_layout(guides = "collect", ncol = 2) + plot_annotation(tag_levels = "A", tag_suffix = ".")
```

@@

The number of iterations necessary to obtain approximately unbiased, confidence-valid estimates corresponds to the $\widehat{R}$ threshold 1.2. The thresholds 1.1 and 1.01 seem too strict compared to the validity of the inferences. The threshold to diagnose significant $AC$-values does not appear to be appropriate in at least the current set-up, and perhaps in iterative imputation in general. A better heuristic to diagnose non-stationarity with $AC$ may be through evaluation of the $AC$-values across iterations: If the values do not substantially decrease with $T$, approximate stationarity may be concluded.
<!-- **[similar to the elbow heuristic in PCA]**. -->

@@

1. First, check traceplots of the default $\theta$s (e.g., chain means and chain variances) for signs of pathological non-convergence. Adjust the imputation model if necessary.

2. Subsequently decide which multivariate $\theta$s to track across iterations. The novel $\theta$ that we propose in this study, $\lambda_1$ is scientific model-independent and may thus always be employed. Alternatively, specify your own scalar summary of interest [see e.g., @buur18]. Monitor these $\theta$s through visual inspection, or using a non-convergence diagnostic.

3. Compute non-convergence diagnostics $\widehat{R}$ and $AC$. Do not use the original implementation to calculate $\widehat{R}$ [@gelm92], or the `R` function `stats::acf()` to compute $AC$ [@R]. Instead, calculate $\widehat{R}$ conform @veht19 and compute autocorrelations manually (see e.g., [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/)).
